---
title: "Results"
editor: visual
---

## Bridge Condition by System

```{r, echo=FALSE, warning=FALSE, message=FALSE}
library(ggplot2)
library(dplyr)
library(tidyr)
data <- data.frame(
  System = c("County System", "First Class Cities", "State System", "NHS State System", 
             "NHS Local System", "Nebraska"),
  Good = c(46.7, 55.0, 70.7, 67.5, 29.7, 58.3),
  Fair = c(34.7, 39.5, 26.2, 29.8, 62.5, 33.9),
  Poor = c(8.3, 4.1, 2.9, 2.3, 4.7, 6.9),
  Closed = c(1.5, 0.6, 0.0, 0.1, 3.1, 1.1)
)

data_long <- data |>
  pivot_longer(cols = Good:Closed, names_to = "Condition", values_to = "Percentage")

ggplot(data_long, aes(x = System, y = Percentage, fill = Condition)) +
  geom_bar(stat = "identity") +
  labs(title = "Bridge Condition by System in Nebraska", 
       x = "Bridge System", 
       y = "Percentage of Bridges") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_fill_manual(values = c("Good" = "lightblue", "Fair" = "steelblue", "Poor" = "darkblue", "Closed" = "gray"))

```

This stacked bar chart displays bridge conditions across various systems in Nebraska. The State System had the highest percentage of bridges in the Good condition (70.7%). First Class Cities had the second highest percentage in the Good condition (55%). The County System had the lowest percentage of Good condition bridges (46.7%). Bridges in Poor condition were the most common in the County System (8.3%). These results suggest that County System Bridges are facing the most infrastructure challenges.

## Bridge Traffic vs Condition

```{r, echo=FALSE, warning=FALSE, message=FALSE}
library(ggplot2)
library(dplyr)
library(tidyr)

bridge_data <- data.frame(
  Bridge = c("US75 over J St", "US75 over Stream", "72nd St/FAU 5037 over UPRR", "US75 over Betz Creek", 
             "42nd St/FAU 5057 over UPRR", "42nd St/FAU 5057 over UPRR", "SB-I180/US34 over I80", 
             "NB-I180/US34 over I80", "N370 over Papillion Creek Trib", "US6 over Saddle Creek Rd", 
             "N50 over I80", "N38 over Big Papillion Creek", "US275/N92 over 72nd Street", 
             "N 14th St/FAU 5227 over Oak Creek", "14th St over US6", "WB-US30/US81 over Loup River", 
             "US6 over Up/BNSF RR", "N31 over Park/Papio", "Bell St over UPRR", "US77 over Elm Creek", 
             "36th St/Fas 5061 over Papillion Creek", "Hamiltn St/Fau5066 over US75", "US30 over Fremont Co Drain Ditch", 
             "Norflk Ave/FAU6020 over N Fk Elkhorn River", "I80 over Link 17B & Rd 77"),
  County = c("Douglas", "Sarpy", "Douglas", "Sarpy", "Douglas", "Douglas", "Lancaster", "Lancaster", 
             "Sarpy", "Douglas", "Sarpy", "Douglas", "Douglas", "Lancaster", "Lancaster", "Platte", 
             "Lancaster", "Douglas", "Dodge", "Dodge", "Sarpy", "Douglas", "Dodge", "Madison", "Cheyenne"),
  Year_Built = c(1970, 1988, 1966, 1989, 1960, 1960, 1960, 1960, 1995, 1934, 1958, 1964, 1962, 1968, 1961, 
                 1931, 1936, 1938, 1994, 1954, 1984, 1977, 1970, 1968, 1974),
  Daily_Crossings = c(85640, 58870, 56260, 38095, 38000, 38000, 32795, 32795, 30705, 26220, 26190, 
                      26100, 24360, 21660, 19190, 14395, 11505, 11100, 9570, 9535, 9470, 8730, 8310, 8130, 7800),
  Condition = c("Poor", "Fair", "Fair", "Fair", "Poor", "Poor", "Good", "Good", "Fair", "Fair", "Fair", 
                "Fair", "Fair", "Good", "Fair", "Good", "Fair", "Poor", "Good", "Fair", "Fair", "Fair", 
                "Fair", "Good", "Fair")
)

summary(bridge_data)

bridge_data$Condition <- factor(bridge_data$Condition, 
                                levels = c("Good", "Fair", "Poor", "Closed"),
                                ordered = TRUE)

bridge_data$Condition_Numeric <- recode(bridge_data$Condition, 
                                        "Good" = 1, "Fair" = 2, "Poor" = 3, "Closed" = 4)
ggplot(bridge_data, aes(x = Daily_Crossings, y = Condition_Numeric, 
                        size = Daily_Crossings, color = Condition)) +
  geom_point(alpha = 0.7) +
  labs(title = "Bridge Traffic vs Condition", 
       x = "Daily Crossings", 
       y = "Condition (1 = Good, 2 = Fair, 3 = Poor, 4 = Closed)", 
       color = "Condition") +
  scale_color_manual(values = c("Good" = "lightblue", 
                                "Fair" = "steelblue", 
                                "Poor" = "darkblue", 
                                "Closed" = "grey")) +
  scale_size_continuous(range = c(1, 5)) +
  theme_minimal(base_family = "Arial") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

The relationship between daily crossing and bridge condition was shown in this scatter plot. The data shows that bridges with higher crossing tend to be in worse condition. Poor condition bridges are seen to be associated with higher traffic volumes (US75 over J St with 85,640 daily crossings), while Good condition bridges tend to have lower traffic. However, the data points indicate considerate variablity because there are some Good and Fair conditioned bridges that have high daily crossings.

## Frequency of Bridge Condition by County

```{r, echo=FALSE, warning=FALSE, message=FALSE}
library(ggplot2)
library(dplyr)
library(tidyr)
ggplot(bridge_data, aes(x = County, fill = Condition)) +
  geom_bar(position = "dodge") +
  labs(title = "Frequency of Bridge Condition by County",
       x = "County", y = "Count of Bridges") +
  scale_fill_manual(values = c("Closed" = "gray", "Poor" = "darkblue", "Fair" = "steelblue", "Good" = "lightblue")) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  theme_minimal(base_family = "Arial") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

This bar chart displays the frequency of bridge conditions across different NE counties. Douglas County has the highest count of Fair condition bridges, followed by Sarpy County. The distribution shows the highest frequency of Fair bridges throughout Nebraska. Douglas County also has the highest frequency of Poor condition bridges.

## Average Condition By Year Build

```{r, echo=FALSE, warning=FALSE, message=FALSE}
bridge_data$Year_Built <- factor(bridge_data$Year_Built)
bridge_data$Year_Built <- droplevels(bridge_data$Year_Built)
bridge_data$Year_Built <- as.numeric(as.character(bridge_data$Year_Built))
bridge_data$Condition_Numeric <- as.numeric(bridge_data$Condition_Numeric)

class(bridge_data$Year_Built)


anova_result <- aov(Condition_Numeric ~ Year_Built, data = bridge_data)

```

|            | Df  | Sum Sq | Mean Sq | F Value | Pr(\>F) |
|------------|-----|--------|---------|---------|---------|
| Year Built | 1   | 0.073  | 0.0727  | 0.171   | 0.683   |
| Residuals  | 23  | 9.767  | 0.4247  |         |         |
|            |     |        |         |         |         |

An ANOVA was conducted to determine whether the year a bridge was built had a significant effect on it's condition rating. The results indicated that their was no significant difference in condition based on year built, (F = 0.171, p = 0.683). This suggests that the age of a bridge alone does not predict its current condition and that it is important to consider other factors that may effect bridge condition.

## Risk Level Assessment

```{r, echo=FALSE, warning=FALSE, message=FALSE}

library(knitr)

summary_stats <- summary(bridge_data$Daily_Crossings)
mean_crossings <- mean(bridge_data$Daily_Crossings)
sd_crossings <- sd(bridge_data$Daily_Crossings)

bridge_data <- bridge_data %>%
  mutate(Risk_Level = case_when(
    Daily_Crossings > 32795 ~ "High", 
    Daily_Crossings > 24360 ~ "Moderate",      
    TRUE ~ "Low"                              
  ))

bridge_data <- bridge_data %>%
  mutate(Risk_Level = case_when(
    Daily_Crossings > 32795 ~ "High", 
    Daily_Crossings > 24360 ~ "Moderate",      
    TRUE ~ "Low"                              
  ))
ggplot(bridge_data, aes(x = County, fill = Risk_Level)) + 
  geom_bar(position = "stack") + 
  labs(title = "Bridge Risk Levels by County", 
       x = "County", 
       y = "Number of Bridges") + 
  scale_fill_manual(values = c("High" = "#1D3557", "Moderate" = "#457B9D", "Low" = "#A8DADC")) +  
  theme_minimal(base_size = 15, base_family = "Arial") + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  

bridge_data <- bridge_data %>%
  mutate(Risk_Level = case_when(
    Daily_Crossings > 32795 ~ "High", 
    Daily_Crossings > 24360 ~ "Moderate",
    TRUE ~ "Low"
  ))
risk_table <- bridge_data %>%
  select(Bridge, County, Daily_Crossings, Condition, Risk_Level)

kable(risk_table, caption = "Bridge Risk Level Assessment Based on Daily Crossings", align = 'l')
```

This stacked bar chart analyzes the distribution of risk levels across Nebraska counties. Douglas County has the highest count of high-risk bridges, followed by Sarpy County. The majority of low-risk bridges are spread more evenly across counties. These results reinforce the need for targeted maintenance efforts in high-traffic Omaha areas.

## Implications

These findings suggest that while there is some correlation between traffic volume and condition, other factors like their maintenance history, are likely influencing the conditions of bridges in Nebraska. The State System generally maintains the best conditions, which could be attributed to better funding. On the other hand, counties like Douglas and Sarpy, which have higher numbers of poor or fair-conditioned bridges require attention to ensure safety on.

The data also reveals a concerning trend: bridges in poorer condition are likely to carry higher traffic volumes, which means placing more stress on vulnerable structures. This highlights a serious risk for disruptions in transportation if a collapse were to occur.

In addition to age and traffic, maintenance history should be further considered in future assessments. While an ANOVA found no direct correlation between the year of construction and bridge condition, regular repairs play an important role in the safety of a bridge.
